The Error Estimate for Computing Electrostatic Interaction in the Inhomogeneous and Correlated Molecular Systems: for the Ewald Summation, SPME and Staggered Mesh Ewald.

### 2012-03-21 longRange.hetero/publish/main.tex
backup the first finished version for publish

\title{The Error Estimate of Force Calculation in the Inhomogeneous Molecular Systems}
\author{Han Wang}
\affiliation{LMAM and School of Mathematical
  Sciences, Peking University}
\author{Pingwen Zhang}
% \email{pzhang@pku.edu.cn}
\affiliation{LMAM and School of Mathematical
  Sciences, Peking University}

\begin{abstract}
\end{abstract}

@@

@@

    \sum_{j=1}^N\,q_iq_j \v K(\v r_i - \v r_j),

@@

@@

    q_i^2\,[\v K\ast\rho_{q}]^2 (\v r_i) \\\label{eqn:thm-error}

@@

@@

    q_i\bigg\langle \sum_{j=1}^Nq_j\v K(\v r_i -\v r_j)\bigg\rangle\\

@@

    q_i\int_{\mathbb R^3}\v K(\v r_i - \v r')

@@

    q_i\, [\v K\ast\rho_q](\v r_i),

@@

    q_jq_k\v K(\v r_i - \v r_j)\cdot\v K(\v r_i - \v r_k)\bigg\rangle \\ \nonumber

@@

    q_j^2\vert\v K(\v r - \v r_j)\vert^2

@@

    q_jq_k \v K(\v r - \v r_j)\cdot\v K(\v r - \v r_k)

@@

    \vert\v K(\v r_i - \v r')\vert^2\rho_{q^2}(\v r')\,\d d\v r'

@@

    \v K(\v r_i - \v r')\cdot\v K(\v r_i - \v r'')\,

@@

    \vert\v K(\v r_i - \v r')\vert^2\rho_{q^2}(\v r')\,\d d\v r'

@@

    \bigg[

@@

    \v K(\v r_i - \v r')\rho_{q}(\v r')\,\d d\v r'
    \bigg]^2

@@

Here are some remarks concerning Theorem~\ref{thm:tmp1}:

@@

  The function $\v K(\v r)$ is called \emph{the error force

@@

  By setting $q_i = q_j = 1$, the theorem~\ref{thm:tmp1} directly proves

@@

  The homogeneity error and the inhomogeneity

@@

\subsection{First neighbor approximation of the correlation
error}

In this section, we consider the first neighbor approximation of the
correlation error. Here ``first neighbors'' means the neighboring
atoms that are connected by the chemical bond.  In most molecular
force fields, such bonds are modeled by a rigid connection between two
atoms.  Therefore, the correlation between such an atom pair is very
strong. As an example, we consider a typical three point charge water
model TIP3P~\cite{jorgensen1983comparison}: each hydrogen atom carries
a charge of $q_{\textrm{H}} = +0.417e$ and each oxygen atom carries a charge of
$q_{\textrm{O}} = -0.834e$. Within one molecule, the H-O bond and H-O-H angle are
rigid. 
% In the system of water, the relative position of the oxygen and
% hydrogen atoms within one molecule is fixed. Therefore, according
% to this property, it is possible to improve the quality of the
% error estimate.
Starts from Eqn.~\eqref{eqn:prf-error-step3},
  \vert\v K(\v r_i - \v r')\vert^2\rho_{q^2}(\v r')\,\d d\v r'
  \\
  \v K(\v r_i - \v r')\cdot\v K(\v r_i - \v r'')\,
  \,\d d\v r'\d d\v r'',
where the definition of $\rho^{(2)}(\v r', \v r'')$ is given by
Eqn.~\eqref{eqn:def-rho2}.
% \begin{align}
%   q_{q^2}(\v r', \v r'') =
%   \bigg\langle
%   \sum_{j\neq k}q_j q_k\delta(\v r' - \v r_j)\delta(\v r'' - \v r_k)
%   \bigg\rangle.
% \end{align}
We seperately consider the first neighbor contribution
and other contributions to
the pairwise density $\rho^{(2)}(\v r', \v r'')$.  Denoting
$\Omega_j = \{ \,k\,\vert\,
\textrm{atom $k$ is the first neighbor of atom $j$} \,\}$,
% the set of the first neighbors of the $j$-th atom by $\Omega_j$,
we have:
\begin{align}\label{eqn:split-rho-q2}
  \sum_j\sum_{k\in\Omega_j}q_j q_k\delta(\v r' - \v r_j)\delta(\v r'' - \v r_k)
  \bigg\rangle
  \sum_j\sum_{k\notin\Omega_j}q_j q_k\delta(\v r' - \v r_j)\delta(\v r'' - \v r_k)
  \bigg\rangle.
For any atom $k$ that is not the first neighbor of atom $j$, we
assume that its position is independent with atom $j$, then
second part on the r.h.s. of the Eqn.~\eqref{eqn:split-rho-q2} is
approximately $\frac{N-3}{N}\rho_q(\v r')\rho_q(\v r'')\approx
\rho_q(\v r')\rho_q(\v r'')$.
% If the number of atoms in the system is
% large, this should be a good approximation.
Therefore, the first part
on the r.h.s. of Eqn.~\eqref{eqn:split-rho-q2} is actually the
pairwise charge corrlation function:
\begin{align}\label{eqn:fnc-correlation-water}
  C_{q^2}(\v r', \v r'') =
  \sum_j\sum_{k\in\Omega_j}q_jq_k\delta(\v r' - \v r_j)\delta(\v r'' - \v r_k)
  \bigg\rangle.
% If connected by the chemical bond, the position of the first neighbors
% $k$ is not independent with the atom $j$, so their correlation is
% strong, and should be calculated explicitly to improve the quality of
% the error estimate.
By inserting Eqn.~\eqref{eqn:fnc-correlation-water} into
the correlation error, we have:
  \mathcal E_{\textrm{correlation}}(\v r_i)
  = q_i^2\sum_j\sum_{k\in\Omega_j}q_jq_k
  \v K(\v r_i - \v r_j)\cdot\v K(\v r_i - \v r_k)
  \big\rangle_{\v r_j,\v r_k}
Denoting the index set of the oxygen atoms by $\Omega_{\textrm{O}}$,
and the index set of the hydrogen atoms by $\Omega_{\textrm{H}}$, 
we have:
  \mathcal E_{\textrm{correlation}}(\v r_i)
  q_i^2\sum_{j\in\Omega_{\textrm{O}}}\sum_{k\in\Omega_j}q_jq_k
  \v K(\v r_i - \v r_j)\cdot\v K(\v r_i - \v r_k)
  \big\rangle_{\v r_j,\v r_k} \\ \nonumber
  q_i^2\sum_{j\in\Omega_{\textrm{H}}}\sum_{k\in\Omega_j}q_jq_k
  \v K(\v r_i - \v r_j)\cdot\v K(\v r_i - \v r_k)
  \big\rangle_{\v r_j,\v r_k} \\\nonumber
  \v K(\v r_i - \v r')\cdot
  q_{\textrm{H}}\v K(\v r_i - \v r' + \v s_{\textrm{H}_1})+
  q_{\textrm{H}}\v K(\v r_i - \v r' + \v s_{\textrm{H}_2})
  \big\rangle\,\rho_{\textrm{O}}(\v r')
  \d d\v r' \\\label{eqn:c-error-1}
  & \,+
  \v K(\v r_i - \v r')\cdot
  q_{\textrm{O}}\v K(\v r_i - \v r' + \v s_{\textrm{O}})+
  q_{\textrm{H}}\v K(\v r_i - \v r' + \v s_{\textrm{H}})
  \big\rangle\,\rho_{\textrm{H}}(\v r')
  \d d\v r'
If positions of the three atom within one molecule are
$\v r_{\textrm{H}_1}$, $\v r_{\textrm{H}_2}$ and $\v r_{\textrm{O}}$, then
$\v s_{\textrm{H}_1} = \v r_{\textrm{H}_1} - \v r_{\textrm{O}}$, 
$\v s_{\textrm{H}_2} = \v r_{\textrm{H}_2} - \v r_{\textrm{O}}$,
$\v s_{\textrm{O}} = \v r_{\textrm{O}} - \v r_{\textrm{H}_1}$ and
$\v s_{\textrm{H}} = \v r_{\textrm{H}_2} - \v r_{\textrm{H}_1}$.
The two ensemble averages in Eqn.~\eqref{eqn:c-error-1}
is taken over all possible water directions 
at position $\v r'$. As an approximation, we
assume these averages are  spacially uniform, namely independet with $\v r'$.
This assumption may be  problematic
in the regions where the preference of water  drection is different
with others.
% In the example considered by the present paper in section~\ref{sec:tmp3},
% it is shown to be a good approximation to obtain the estimate of the
% correlation error.
One possible way to calculate the ensemble averages in
Eqn.~\eqref{eqn:c-error-1} is by the Fourier transform
of the error force kernel:
  &\big\langle
  q_{\textrm{H}}\v K(\v r_i - \v r' + \v s_{\textrm{H}_1})+
  q_{\textrm{H}}\v K(\v r_i - \v r' + \v s_{\textrm{H}_2})
  \big\rangle \\ \nonumber
  \frac1V\bigg\langle
  \sum_{\v m}
  q_{\textrm{H}}\hat{\v K}(\v m)
  e^{2\pi i\v m\cdot(\v r_i - \v r' + \v s_{\textrm{H}_1})} +
  \sum_{\v m}
  q_{\textrm{H}}\hat{\v K}(\v m)
  e^{2\pi i\v m\cdot(\v r_i - \v r' + \v s_{\textrm{H}_2})}
  \bigg\rangle \\\nonumber
  \frac1V\sum_{\v m}
  \hat{\v K}(\v m)
  e^{2\pi i\v m\cdot(\v r_i - \v r')}
  \,  
  \langle
  q_{\textrm{H}}\,e^{2\pi i\v m\cdot\v s_{\textrm{H}_1}} +
  q_{\textrm{H}}\,e^{2\pi i\v m\cdot\v s_{\textrm{H}_2}}
  \rangle,
where ``$\wedge$'' denotes the forward Fourier transform. By denoting
  \hat T_{\textrm{O}}(\v m)
  \langle
  q_{\textrm{H}}\,e^{2\pi i\v m\cdot\v s_{\textrm{H}_1}} +
  q_{\textrm{H}}\,e^{2\pi i\v m\cdot\v s_{\textrm{H}_2}}
  \rangle,\\
  \hat T_{\textrm{H}}(\v m)
  \langle
  q_{\textrm{O}}\,e^{2\pi i\v m\cdot\v s_{\textrm{O}}} +
  q_{\textrm{H}}\,e^{2\pi i\v m\cdot\v s_{\textrm{H}}}
  \rangle.
The Eqn.~\eqref{eqn:c-error-1} becomes:
  \mathcal E_{\textrm{correlation}}(\v r_i)
  =&\,
  \v K(\v r_i - \v r')\cdot
  (\hat T_{\textrm{O}}\hat{\v K})^{\vee}(\v r_i - \v r')
  \,\rho_{\textrm{O}}(\v r')
  \,\d d\v r' \\\label{eqn:c-error-water}
  & \,+
  \v K(\v r_i - \v r')\cdot
  (\hat T_{\textrm{H}}\hat{\v K})^{\vee}(\v r_i - \v r')
  \,\rho_{\textrm{H}}(\v r')
  \,\d d\v r',
where ``$\vee$'' denotes the backward Fourier transform.
In the Fourier space, the error force kernel $\v K(\v r)$ is
influenced by the prefactors $T_{\textrm{O}}$ and $T_{\textrm{H}}$,
which are accounting for the atomic correlation within one water
molecule.

@@

Both the direct part and reciprocal part of the Ewald summation can be
written in the pairwise form of Eqn.~\eqref{eqn:thm-error-force}.  The direct
error force kernel is defined by:
% The direct part is a short-range interaction in real space, so it is
% usually treated by the cut-off method, which gives a error force
% kernel of
\begin{align}
  \v K^{\textrm{cut}}_{\textrm{dir}}(\v r) =
  \left\{
    \begin{aligned}
      &\,0, &\quad & r \leq r_c;\\
      &
      \Big[
      \frac{2\beta}{\sqrt\pi} e^{-\beta^2r^2} + \frac{\textrm{erfc}(\beta r)}{r}
      \Big]\frac{\v r}{r^2}
      ,& & r > r_c,
      % \frac{\textrm{erfc}(\beta\vert\v r\vert)}{\vert\v r\vert}
    \end{aligned}
  \right.
\end{align}
where $r = \vert\v r\vert$, and $r_c$ is the cut-off radius in the
real space.
% The reciprocal error force, namely the difference between
% the the exact and truncated reciprocal force, is
% \begin{align}
%   \Delta \v F^{\textrm{tr}}_{\textrm{rec}}(\v r_i)
%   = & - 
%   \sum_j\,q_iq_j
%   \sum_{\vert m_\alpha\vert \geq K_\alpha/2}
%   \v g(\v m) \,
%   e^{2\pi i\v m\cdot\v r_{ij}}.
% \end{align}
% Therefore,
The reciprocal error force kernel is given by:
\begin{align}
  \v K^{\textrm{tr}}_{\textrm{rec}}(\v r) =
  \sum_{
      \vert m_\alpha\vert \geq K_\alpha/2}
  \v g(\v m) \,
  e^{2\pi i\v m\cdot\v r}.
\end{align}
By the Therefore \ref{thm:tmp1}, the error estimate of the
truncated Ewald summation is straightforward.

@@

error force of ik-differentiation~\eqref{eqn:ik.error.force} cannot be written
in the form of~\eqref{eqn:thm-error-force}, so the error estimate derived by
Theorem~\ref{thm:tmp1} cannot be used here. However, it is still
possible to develop the error estimate for the ik-differentiation.

@@

  \bigg] (\v r_i),

@@

  \bigg[\,

@@

  \,\bigg] (\v r_i)\\\nonumber

@@

  \bigg] (\v r_i) \\\label{eqn:ik.error}

@@

@@

\newpage
\appendix
\section{Error estimate of the Ewald summation}
\label{sec:appendix}

>>>>

\title{The Error Estimate for Computing Electrostatic Interaction
  in the Inhomogeneous and Correlated Molecular Systems:
  for the Ewald Summation, SPME and Staggered Mesh Ewald.
}
\author{Han Wang}
\email{han.wang@fu-berlin.de}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Christof Sch\"utte}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Pingwen Zhang}
% \email{pzhang@pku.edu.cn}
\affiliation{LMAM and School of Mathematical
  Sciences, Peking University}

\begin{abstract}

\end{abstract}

@@

\section{Introduction}

The calculation of long-range electrostatic interaction is a very
important topic in the molecular simulations.  Due to the very slow
decay of the interaction with respect to the distance between charges,
the cut-off and neighbor list algoritm does not work, and moreover the
energy of the system is only conditionally convergent.  One of the
first algoritms handling this problem is the Ewald summation that
dates back to 1920's~\cite{ewald1921die}.  However, it is shown that
with a careful tuning of working parameters, the optimal computational
cost of the Ewald summation is of $\mathcal
O(N^{3/2})$~\cite{perram1988asc}, which is not feasible for morden
large scale molecular simulations. The particle mesh Ewald (PME)
method~\cite{darden1993pme}, the smooth particle mesh Ewald (SPME)
method~\cite{essmann1995spm} and the particle-particle-particle-mesh
(P3M) method~\cite{hockney1988computer, deserno1998mue1} are the most
popular Ewald-based fast algorithms that reduces the computational
cost to $\mathcal O(N\log N)$. More recently, the staggered mesh or
``interlacing'' technique firtly developed in
1970's~\cite{chen1974reduction,eastwood1976optimal} is applied to SPME
(called staggered mesh Ewald)~\cite{cerutti2009staggered} and P3M
(called interlaced P3M)~\cite{neelov2010interlaced}, and is showed to
improve the accuracy of these algorithms.

The fast algoritms have achieved great success in the computer
simulation of charged molecular systems over the past three decades.
However, the fact that the user should provide six working parameters
for these methods may be problematic: it is diffcult to try out the
most efficient parameter set, and an arbitary choice may lead to
qualitively wrong simulation result. This problem can be solved by an
in-depth study of the error introduced by these algrotims, namly the
error estimate~\cite{petersen1995accuracy, deserno1998mue2,
  wang2010optimizing, neelov2010interlaced, ballenegger2012convert},
by which the error is described by a function of the working
parameter.  By these error estimates, a work flow was designed to
automatically determine the nearly optimal combination of parameters,
in the sense that it is fastest one given a demanding
accuracy~\cite{wang2010optimizing}. However, so far, all error
estimates assume the homogeneity of the charge distribution, and the
independency of any pair of charges (i.e. the charges are
uncorrelated), but neither of them is satisfactorily fulfilled in most
realistic systems of practical interest.  For example, the lipid biomembrane
solving in water: the carbon chains of the lipids carry no charge, and
the charges are correlated by covalent bonds, van der Waals
interaction, hydrogen bonding network, etc..  Therefore, the current
error estimates may be problematic, or even not applicable.  A lot of
research have shown that due to the correlation of charges, the error estimates
obviously deviate from the real
error~\cite{deserno1998mue2,wang2010optimizing,ballenegger2012convert}.

The purpose of this research is to develop an realiable force error
estimate in inhomogeneous and strongly correlated molecular systems.
We estimate the force error because the application is the molecular
dynamics (MD) simulation.  Firstly, a most general error estimate
frame work for the pairwise interactions is set up. We prove that the
error is decomposed into three additive parts: the homogeneity error,
the inhomogeneity error and the correlation error~\cite{wang2012}. The
homogeneity and inhomogeneity error can be calculated at a cost of
$\mathcal O(N\log N)$ given the convolutional form of the error
estimates.  Then a way is proposed to calculate the approximation of
correlation error at cost of $\mathcal O(N\log N)$. Though it is
possible to improve the accuracy of the approximation systematically,
numerical results show that it is good enough
to consider the correlation stems from
the neighboring atoms within the same molecule
in an inhomogeneous water system.  Following this framework, we
develop the error estimate for the Ewald summation, the SPME method
and the staggered method Ewald method. We study the SPME method
because it is the one of the most popular long-range algorithms among
the mainstream MD simulation packages, for example,
AMBER~\cite{case2005amber}, GROMACS~\cite{van2005gromacs,
  hess2008gromacs} and NAMD~\cite{phillips2005scalable}.  The estimate
of the homogeneity and inhomogeneous errors is tested by by two
randomly distributed inhomogeneous charges systems, and is proved to
be very sharp in these ideal systems.  By studying a real
inhomogeneous water system, the nearest neigbor approximation of
correlation error is proved to improve the quality of error estimate
greatly. This paper is ended by conclusions and remarks in
section~\ref{sec:conclusions}.

% Then a
% systematical approach to calculate the correlation error app

@@

    \sum_{j=1}^N\,q_iq_j \v K(\v r_i, \v r_j),

@@

    =q_i\, \int_{\mathbb R^3}\v K (\v r_i,\v r')\,\rho_q(\v r')\,\d d\v r',

@@

    q_i^2\,\int_{\mathbb R^3}\vert\v K(\v r_i,\v r')\vert^2\,\rho_{q^2} (\v r')\,\d d\v r' + 
    q_i^2\,
    \int_{\mathbb R^3}\v K(\v r_i,\v r')\,\rho_{q}(\v r')\,\d d\v r'
    \,\big]^2
    \\\label{eqn:thm-error}

@@

    \v K(\v r_i,\v r')\cdot
    \v K(\v r_i,\v r'')\,

@@

    q_i\bigg\langle \sum_{j=1}^Nq_j\v K(\v r_i,\v r_j)\bigg\rangle\\

@@

    q_i\int_{\mathbb R^3}\v K(\v r_i, \v r')

@@

    q_i\,\int\v K(\v r_i,\v r')\,\rho_q(\v r')\,\d d\v r'.

@@

    q_jq_k\v K(\v r_i, \v r_j)\cdot\v K(\v r_i, \v r_k)\bigg\rangle \\ \nonumber

@@

    q_j^2\vert\v K(\v r, \v r_j)\vert^2

@@

    q_jq_k \v K(\v r, \v r_j)\cdot\v K(\v r, \v r_k)

@@

    \vert\v K(\v r_i, \v r')\vert^2\rho_{q^2}(\v r')\,\d d\v r'

@@

    \v K(\v r_i , \v r')\cdot\v K(\v r_i , \v r'')\,

@@

    \vert\v K(\v r_i , \v r')\vert^2\rho_{q^2}(\v r')\,\d d\v r'

@@

@@

    \v K(\v r_i , \v r')\rho_{q}(\v r')\,\d d\v r'
    \big]^2

@@

\begin{corollary}\label{thm:tmp2}
  If the function $\v K$ has the property
    \label{eqn:cor-kernel}
    \v K(\v r_i,\v r_j) = \v K(\v r_i - \v r_j),    
  then the error estimates are in the convolution form, namely:
  \begin{align}\label{eqn:cor-meanf}
  and 
    q_i^2\,[\v K\ast\rho_{q}]^2 (\v r_i) \\\label{eqn:cor-error}
  Where $\ast$ denotes the convolution.
\end{corollary}

Here are some remarks concerning Theorem~\ref{thm:tmp1} and Corollary~\ref{thm:tmp2}:

@@

  The function $\v K(\v r_i, \v r_j)$ is called \emph{the error force

@@

  By setting $q_i = q_j = 1$, the corollary~\ref{thm:tmp2} directly proves

@@

  Without the property $\v K(\v r_i,\v r_j) = \v K(\v r_i - \v r_j)$,
  theorem~\ref{thm:tmp1} calculate the homogeneity and inhomogeneity
  error at a cost of $\mathcal O(N^2\log N)$, and correlation error
  at a cost of $\mathcal O(N^3\log N)$.
  By writting the error estimates in the covolution form in
  the corollary~\ref{thm:tmp2},
  the homogeneity error and the inhomogeneity

@@

\subsection{Nearest neighbor approximation of the correlation
error}

In this section, we consider the nearest neighbor approximation of the
correlation error. Here the definition of the ``nearest neighbor'' is
very flexible, basically it means the neigbors of a charge that are
not very far away. It could be defined by the charges falling in some
certain neighboring range or charge connected by a chemical bonds.
% Here ``first neighbors'' means the neighboring
% atoms that are connected by the chemical bond.  In most molecular
% force fields, such bonds are modeled by a rigid connection between two
% atoms.  Therefore, the correlation between such an atom pair is very
% strong.
% In the system of water, the relative position of the oxygen and
% hydrogen atoms within one molecule is fixed. Therefore, according
% to this property, it is possible to improve the quality of the
% error estimate.
Starts from Eqn.~\eqref{eqn:prf-error-step3},
\begin{align} \nonumber
  \langle\vert\Delta\v F(\v r_i)\vert^2\rangle
  =\,&
  q_i^2\int_{\mathbb R^3}
  \vert\v K(\v r_i , \v r')\vert^2\rho_{q^2}(\v r')\,\d d\v r'
  \\
  & +
  q_i^2\int_{\mathbb R^3\times\mathbb R^3}
  \v K(\v r_i , \v r')\cdot\v K(\v r_i , \v r'')\,
  \rho^{(2)}(\v r', \v r'')
  \,\d d\v r'\d d\v r'',
\end{align}
where the definition of $\rho^{(2)}(\v r', \v r'')$ is given by
Eqn.~\eqref{eqn:def-rho2}.
% \begin{align}
%   q_{q^2}(\v r', \v r'') =
%   \bigg\langle
%   \sum_{j\neq k}q_j q_k\delta(\v r' - \v r_j)\delta(\v r'' - \v r_k)
%   \bigg\rangle.
% \end{align}
We seperately consider the  contribution from nearest neigbors
and other contributions to
the pairwise density $\rho^{(2)}(\v r', \v r'')$.  Denoting
$\Omega_j = \{ \,k\,\vert\,
\textrm{atom $k$ is the nearest neighbor of atom $j$} \,\}$,
% the set of the first neighbors of the $j$-th atom by $\Omega_j$,
we have:
\begin{align}\label{eqn:split-rho-q2}
  \rho^{(2)}(\v r', \v r'')
  = &\,
  \bigg\langle
  \sum_j\sum_{k\in\Omega_j}q_j q_k\delta(\v r' - \v r_j)\delta(\v r'' - \v r_k)
  \bigg\rangle
  +
  \bigg\langle
  \sum_j\sum_{k\notin\Omega_j}q_j q_k\delta(\v r' - \v r_j)\delta(\v r'' - \v r_k)
  \bigg\rangle.
\end{align}
We assume two charges are independent if they are not nearest neighbors.
Then
second part on the r.h.s. of the Eqn.~\eqref{eqn:split-rho-q2} is
approximately $\frac{N-\#(\Omega_j)}{N}\rho_q(\v r')\rho_q(\v r'')\approx
\rho_q(\v r')\rho_q(\v r'')$, where $\#(\Omega_j)$ is the number of nearest neighbors considered. This approximation is valid only when the number
of neigbors considered is small comparing with the number of charges
in the system.
% If the number of atoms in the system is
% large, this should be a good approximation.
Therefore, the first part
on the r.h.s. of Eqn.~\eqref{eqn:split-rho-q2} is actually the
pairwise charge corrlation function:
\begin{align}\label{eqn:fnc-correlation-water}
  C_{q^2}(\v r', \v r'') \approx
  \bigg\langle
  \sum_j\sum_{k\in\Omega_j}q_jq_k\delta(\v r' - \v r_j)\delta(\v r'' - \v r_k)
  \bigg\rangle.
\end{align}
% If connected by the chemical bond, the position of the first neighbors
% $k$ is not independent with the atom $j$, so their correlation is
% strong, and should be calculated explicitly to improve the quality of
% the error estimate.
By inserting Eqn.~\eqref{eqn:fnc-correlation-water} into
the correlation error, we have:
\begin{align}
  \mathcal E_{\textrm{correlation}}(\v r_i)
  \approx q_i^2\sum_j\sum_{k\in\Omega_j}q_jq_k
  \big\langle
  \v K(\v r_i , \v r_j)\cdot\v K(\v r_i , \v r_k)
  \big\rangle_{\v r_j,\v r_k}
\end{align}
The quality of this approximation depends on how to define the
``nearest neighbors''. In the system of liquid and gas, the
correlation is only short-ranged, so it is possible to systematically
impove the accuracy by considering more and more neigbors as long as
the number of neigbors is small comparing with the size of the system.

As an example, we consider a typical three point charge water model
TIP3P~\cite{jorgensen1983comparison}: each hydrogen atom carries a
charge of $q_{\textrm{H}} = +0.417e$ and each oxygen atom carries a
charge of $q_{\textrm{O}} = -0.834e$. Within one molecule, the H-O
bond and H-O-H angle are rigid, so the correlation within one molecule
is very strong.  Therefore, in this case, we define the nearest neighbor by the
atoms within the same water molecule.  Denoting the index set of the oxygen
atoms by $\Omega_{\textrm{O}}$, and the index set of the hydrogen
atoms by $\Omega_{\textrm{H}}$, we have:
\begin{align}\nonumber
  \mathcal E_{\textrm{correlation}}(\v r_i)
  = &\,
  q_i^2\sum_{j\in\Omega_{\textrm{O}}}\sum_{k\in\Omega_j}q_jq_k
  \big\langle
  \v K(\v r_i , \v r_j)\cdot\v K(\v r_i , \v r_k)
  \big\rangle_{\v r_j,\v r_k} \\ \nonumber
  &\,+ 
  q_i^2\sum_{j\in\Omega_{\textrm{H}}}\sum_{k\in\Omega_j}q_jq_k
  \big\langle
  \v K(\v r_i , \v r_j)\cdot\v K(\v r_i , \v r_k)
  \big\rangle_{\v r_j,\v r_k} \\\nonumber
  = &\,
  q_i^2
  \int_{\mathbb R^3}
  \v K(\v r_i , \v r')\cdot
  2\big\langle
  q_{\textrm{H}}\v K(\v r_i , \v r' + \v s_{\textrm{O}})
  \big\rangle\,
  \rho_{\textrm{O}}(\v r')
  \d d\v r' \\\label{eqn:c-error-1}
  & \,+
  q_i^2
  \int_{\mathbb R^3}
  \v K(\v r_i , \v r')\cdot
  \big\langle
  q_{\textrm{O}}\v K(\v r_i , \v r' - \v s_{\textrm{O}})+
  q_{\textrm{H}}\v K(\v r_i , \v r' + \v s_{\textrm{H}})
  \big\rangle\,\rho_{\textrm{H}}(\v r')
  \d d\v r'
\end{align}
If positions of the three atom within one molecule are
$\v r_{\textrm{H}_1}$, $\v r_{\textrm{H}_2}$ and $\v r_{\textrm{O}}$, then
% $\v s_{\textrm{H}_1} = \v r_{\textrm{H}_1} - \v r_{\textrm{O}}$, 
% $\v s_{\textrm{H}_2} = \v r_{\textrm{H}_2} - \v r_{\textrm{O}}$,
% $\v s_{\textrm{O}} = \v r_{\textrm{O}} - \v r_{\textrm{H}_1}$ and
% $\v s_{\textrm{H}} = \v r_{\textrm{H}_2} - \v r_{\textrm{H}_1}$.
$\v s_{\textrm{O}} = \v r_{\textrm{H}_1} - \v r_{\textrm{O}}$ and
$\v s_{\textrm{H}} = \v r_{\textrm{H}_2} - \v r_{\textrm{H}_1}$.
The two ensemble averages in Eqn.~\eqref{eqn:c-error-1}
is taken over all possible water directions 
at position $\v r'$. As an approximation, we
assume these averages are  spacially uniform, namely independet with $\v r'$.
This assumption may be  problematic
in the regions where the preference of water  drection is different
with others.

To calculate the nearest neighbor approximated correlation error,
we assume the error force kernel
has the form $\v K(\v r_i , \v r_j) = \v K(\v r_i - \v r_j)$, so
the integrals can be written in the covolution form and accelerated by the FFT.
% In the example considered by the present paper in section~\ref{sec:tmp3},
% it is shown to be a good approximation to obtain the estimate of the
% correlation error.
One possible way to calculate the ensemble averages in
Eqn.~\eqref{eqn:c-error-1} is by the Fourier transform
of the error force kernel:
\begin{align} \nonumber
  \big\langle
  q_{\textrm{H}}\v K(\v r_i - \v r' + \v s_{\textrm{O}})
  \big\rangle 
  =\,&
  \frac1V\bigg\langle
  \sum_{\v m}
  q_{\textrm{H}}\hat{\v K}(\v m)
  e^{2\pi i\v m\cdot(\v r_i - \v r' + \v s_{\textrm{O}})} 
  \bigg\rangle \\\nonumber
  =\,&
  \frac1V\sum_{\v m}
  \langle
  q_{\textrm{H}}\,e^{2\pi i\v m\cdot\v s_{\textrm{O}}}
  \rangle\,
  \hat{\v K}(\v m)\,
  e^{2\pi i\v m\cdot(\v r_i - \v r')},
\end{align}
where ``$\wedge$'' denotes the forward Fourier transform. By denoting
\begin{align}
  \hat T_{\textrm{O}}(\v m)
  &= 
  2\langle
  q_{\textrm{H}}\,e^{2\pi i\v m\cdot\v s_{\textrm{O}}}
  \rangle,\\
  \hat T_{\textrm{H}}(\v m)
  &= 
  \langle
  q_{\textrm{O}}\,e^{-2\pi i\v m\cdot\v s_{\textrm{O}}} +
  q_{\textrm{H}}\,e^{2\pi i\v m\cdot\v s_{\textrm{H}}}
  \rangle.
\end{align}
The Eqn.~\eqref{eqn:c-error-1} becomes:
\begin{align}\nonumber
  \mathcal E_{\textrm{correlation}}(\v r_i)
  =&\,
  q_i^2
  \int_{\mathbb R^3}
  \v K(\v r_i - \v r')\cdot
  (\hat T_{\textrm{O}}\hat{\v K})^{\vee}(\v r_i - \v r')
  \,\rho_{\textrm{O}}(\v r')
  \,\d d\v r' \\\label{eqn:c-error-water}
  & \,+
  q_i^2
  \int_{\mathbb R^3}
  \v K(\v r_i - \v r')\cdot
  (\hat T_{\textrm{H}}\hat{\v K})^{\vee}(\v r_i - \v r')
  \,\rho_{\textrm{H}}(\v r')
  \,\d d\v r',
\end{align}
where ``$\vee$'' denotes the backward Fourier transform.
In the Fourier space, the error force kernel $\v K(\v r)$ is
influenced by the prefactors $T_{\textrm{O}}$ and $T_{\textrm{H}}$,
which are accounting for the atomic correlation within one water
molecule.

@@

Both the error force kernel of the direct and reciprocal
part of the Ewald summation can be
written in the form of Eqn.~\eqref{eqn:cor-kernel}.  The direct
part is:
% The direct part is a short-range interaction in real space, so it is
% usually treated by the cut-off method, which gives a error force
% kernel of
\begin{align}
  \v K^{\textrm{cut}}_{\textrm{dir}}(\v r) =
  \left\{
    \begin{aligned}
      &\,0, &\quad & r \leq r_c;\\
      &
      \Big[
      \frac{2\beta}{\sqrt\pi} e^{-\beta^2r^2} + \frac{\textrm{erfc}(\beta r)}{r}
      \Big]\frac{\v r}{r^2}
      ,& & r > r_c,
      % \frac{\textrm{erfc}(\beta\vert\v r\vert)}{\vert\v r\vert}
    \end{aligned}
  \right.
\end{align}
where $r = \vert\v r\vert$, and $r_c$ is the cut-off radius in the
real space.
% The reciprocal error force, namely the difference between
% the the exact and truncated reciprocal force, is
% \begin{align}
%   \Delta \v F^{\textrm{tr}}_{\textrm{rec}}(\v r_i)
%   = & - 
%   \sum_j\,q_iq_j
%   \sum_{\vert m_\alpha\vert \geq K_\alpha/2}
%   \v g(\v m) \,
%   e^{2\pi i\v m\cdot\v r_{ij}}.
% \end{align}
% Therefore,
The reciprocal error force kernel is given by:
\begin{align}
  \v K^{\textrm{tr}}_{\textrm{rec}}(\v r) =
  \sum_{
      \vert m_\alpha\vert \geq K_\alpha/2}
  \v g(\v m) \,
  e^{2\pi i\v m\cdot\v r}.
\end{align}
By the corollary~\ref{thm:tmp2}, the error estimate of the
truncated Ewald summation is straightforward.

@@

error force kernel of ik-differentiation
cannot be written
in the form of~\eqref{eqn:cor-kernel}, so the error estimate derived by
corollary~\ref{thm:tmp2} cannot be used here. However, it is still
possible to develop the error estimate with convolution form.

@@

  \big] (\v r_i),

@@

@@

  \,\big] (\v r_i)\\\nonumber

@@

  \big] (\v r_i) \\\label{eqn:ik.error}

@@

The nearest neighbor approximation to the correlation error is:
  \mathcal E^{\textrm{ik}}_{\textrm{correlation}}(\v r_i)
  \big[\,
  \v G_{\alpha,l}\cdot[\hat T_{\textrm{O}}\hat{\v G}_{\alpha,l}]^{\vee}
  \ast \rho_{\textrm{O}}
  \,\big] (\v r_i) \\\nonumber
  \,&+
  \cdot
  [\,\hat T_{\textrm{O}}\hat{\v G}_{\alpha,l}]^{\vee}
  \ast \rho_{\textrm{O}}\,
  \big] (\v r_i) \\\nonumber
  \,&+
  \big[\,
  \v G_{\alpha,l}\cdot[\hat T_{\textrm{H}}\hat{\v G}_{\alpha,l}]^{\vee}
  \ast \rho_{\textrm{H}}
  \,\big] (\v r_i) \\
  \,&+
  \cdot
  [\,\hat T_{\textrm{H}}\hat{\v G}_{\alpha,l}]^{\vee}
  \ast \rho_{\textrm{H}}\,
  \big] (\v r_i).

@@

\newpage
\appendix
\section{Error estimate of the smooth particle mesh
Ewald method}
\label{sec:appendix}
